---
title: "ETC5513-Assignment4"
author:
- familyname: Moody
  othernames: Lachlan
  address: Monash University
  email: ltmoo2@student.monash.edu.au
  correspondingauthor: true
  qualifications: Student:27809951
- familyname: 
  othernames: 
  address: Monash University
  email: 
  correspondingauthor: true
  qualifications:
- familyname: 
  othernames: 
  address: Monash University
  email: 
  correspondingauthor: true
  qualifications:
- familyname: 
  othernames: 
  address: Monash University
  email: 
  correspondingauthor: true
  qualifications:
department: Department of\newline Econometrics &\newline Business Statistics
organization: Acme Corporation
bibliography: references.bib
biblio-style: authoryear-comp
linestretch: 1.5
output:
  MonashEBSTemplates::report:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, messages=FALSE, warning=FALSE, fig.pos = 'H')

library(tidyverse)
library(ggplot2)
library(kableExtra)
library(here)
library(ggthemes)
library(countrycode)
library(gganimate)
library(broom)
library(knitr)

```


```{r readingdata, message = FALSE, warning = FALSE}
expectancy <- read_csv(here::here("data", "life_expectancy.csv"))
names(expectancy) <- paste(names(expectancy), expectancy[1, ], sep = "_") ## Merging first two header rows together to get row names
expectancy <- expectancy[-1,] ## Removing now unnecessary firt row

spend <- read_csv(here::here("data", "health_expenditure.csv"), skip = 1)

gdp <- read_csv(here::here("data", "gdp_data.csv"), skip = 3)

head(expectancy,10)
head(spend, 10)
head(gdp, 10)
```

```{r}
expectancy_clean <- expectancy %>%
  pivot_longer(cols = `Life expectancy at birth (years)_Both sexes`:`Healthy life expectancy (HALE) at age 60 (years)_2_Female`,
               names_to = c("Type", "Gender"),
               names_sep = "_",
               values_to = "Expectancy") %>%
  mutate(Gender = if_else(Gender == "1", "Male",
                          if_else(Gender == 2, "Female", "Both"))) %>%
  rename(Country = X1_Country,
         Year = X2_Year) %>%
  mutate(Expectancy = as.numeric(Expectancy)) %>%
  mutate(Year = as.numeric(Year))
```



\section*{Introduction}



\section*{Differences between men and women}

The disparity in the life expectancy between men and women was first recognised in the 1920's [@luy2014gender]. This section will examine this difference in detail.

The first step was to tidy the life expectancy data into a format R. The original variables were condensed into Country, Year, Type (expectancy measure), Gender and Expectancy. The first 10 resulting rows are displayed in Table \@ref(tab:transform). 

```{r transform}
expectancy_clean <- expectancy %>%
  pivot_longer(cols = `Life expectancy at birth (years)_Both sexes`:`Healthy life expectancy (HALE) at age 60 (years)_2_Female`,
               names_to = c("Type", "Gender"),
               names_sep = "_",
               values_to = "Expectancy") %>%
  mutate(Gender = if_else(Gender == "1", "Male",
                          if_else(Gender == 2, "Female", "Both"))) %>%
  rename(Country = X1_Country,
         Year = X2_Year) %>%
  mutate(Expectancy = as.numeric(Expectancy)) %>%
  mutate(Year = as.numeric(Year))

expectancy_clean %>%
  head(10) %>%
  kable(caption = "Life expectancy data") %>%
  kable_styling(bootstrap_options = c("striped", "bordered"))
```

An average was then calculated across the recorded gender types for life expectancy at birth which is shown in Table \@ref(tab:average). It appears that the average life expectancy for both genders is slightly over 69 years with females having the higher value of 71.6 years and males almost 5 years lower at 66.9 years.

```{r average}
averages <- expectancy_clean %>%
  filter(str_detect(Type, "Life expectancy at birth")) %>%
  group_by(Gender) %>%
  summarise("Average Life Expectancy" = mean(Expectancy, na.rm = TRUE))

averages %>%
  kable(caption = "Average life expectancy across genders", digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"))
```

The yearly averages were then visualised using the boxplot below in Figure \@ref(fig:box) with gender on the x-axis and the average life expectancy on the y-axis. This further illustrates that females on average have a higher life expectancy across all years in the data set, having a higher median and interquartile range than men with only the male maximum and female minimum values overlapping.

```{r box, fig.cap = "Comparison of life expectancy across genders per year"}

yearly_avg <- expectancy_clean %>%
  filter(str_detect(Type, "Life expectancy at birth")) %>%
  group_by(Gender, Year) %>%
  summarise(Average = mean(Expectancy, na.rm = TRUE))

yearly_avg %>%
  ggplot(aes(x = Gender,
             y = Average,
             colour = Gender)) +
  geom_boxplot() +
  theme_bw() +
  ylab("Average life expectancy")
```

A line graph was then produced with year on the x-axis and the average life expectancy on the y-axis to illustrate this trend. As can be seen in Figure \@ref(fig:test), across all years recorded in the data set, females have had a higher life expectancy than males by approximately the same amount.

```{r test, fig.cap = "Change in life expectancy by year for each gender"}

yearly_avg %>%
  ggplot(aes(x = Year,
             y = Average,
             group = Gender,
             colour = Gender)) +
  geom_line(size = 1) +
  ylab("Average Life Expectancy") +
  theme_bw()

```

A linear model was then produced to see if this trend was due to gender or year. The outputs are shown in Table \@ref(tab:linear). 

```{r linear}
library(broom)

reg_data <- yearly_avg %>%
  filter(Gender != "Both") %>%
  mutate(Year = Year-2000)

gender_reg <- lm(Average ~ Gender,
   data = reg_data)

exp_reg <-  lm(Average ~ Year,
   data = reg_data)

both_reg <- lm(Average ~ Gender + Year,
   data = reg_data)

gender_r <- glance(gender_reg) %>%
  select(r.squared, adj.r.squared) %>%
  mutate(Model = "Gender")

year_r <- glance(exp_reg) %>%
  select(r.squared, adj.r.squared) %>%
  mutate(Model = "Year")

both_r <- glance(both_reg) %>%
  select(r.squared, adj.r.squared) %>%
  mutate(Model = "Both")

reg_comp <- bind_rows(Gender = tidy(gender_reg),
          Year = tidy(exp_reg),
          Both = tidy(both_reg),
          .id = "Model") %>%
              left_join(bind_rows(gender_r, 
                      year_r, both_r),
            by = "Model") %>%
  select(Model,
         term, 
         estimate, 
         r.squared, 
         adj.r.squared)

reg_comp %>%
  kable(caption = "Model estimates for life expectancy based on gender and year", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"))

```

Several insights are provided from this table. Firstly, the GenderMale term for the Gender model suggests that males on average live 4.7 years less than the female average of 71.6. Secondly, the Year term for the Year model shows that for each after 2000, people on average are living 0.35 years longer. Finally the adjusted.r.squared values describes the explanatory power of these models. The results have been plotted in Figure \@ref(fig:modelcomp) with the model name on the x-axis and the adjusted r squared on the y-axis.

```{r modelcomp, fig.cap = "Comparison of model fits"}
reg_comp %>%
  ggplot(aes(x = Model,
             y = adj.r.squared,
             colour = Model)) +
  geom_point(size = 3) +
  ylab("Adjusted r squared") +
  theme_bw()
```

This shows that the gender model is superior to the year mode, explaining over 64% of the variation in the life expectancy value compared to under 33% for the year variable. Interestingly, almost the entire variation in life expectancy can be explained by including both variables.

\section*{Life expectancy and Health expenditure}

```{r country, include=FALSE}
country_le <- expectancy_clean %>% 
  filter(Gender == "Both") %>% 
  pivot_wider(names_from = Type, 
  values_from = Expectancy)

#examine data
head(country_le)

```


```{r continent}
# Add continent variable
country_le_continent <- country_le %>% 
  mutate(continent = countrycode(sourcevar = country_le$Country, 
                                            origin = "country.name",
                                            destination = "continent")) %>% 
  mutate(Year = as.character(Year))
  
```


```{r tidy_data}
# tidy health expenditure data
health_expenditure <- spend %>% 
  select(-2) %>%   #remove year2017 to match the life expectancy data
  pivot_longer(cols = c(2:18), names_to = "Year", 
               values_to = "Health expenditure per capita(US$)") 
  
```

```{r le_hs}
# join life_expectancy and health_expenditure
life_health_expenditure <- left_join(country_le_continent, 
                               health_expenditure, 
                               by = c("Country" = "Country", 
                                      "Year" = "Year")) %>% 
  mutate(Year = as.integer(Year),
         `Health expenditure per capita(US$)` = as.numeric(`Health expenditure per capita(US$)`),
         `Life expectancy at birth (years)` = as.numeric(`Life expectancy at birth (years)`),
         `Life expectancy at age 60 (years)` = as.numeric(`Life expectancy at age 60 (years)`))

```

## Dynamic plots for Life expectancy and Health expenditure

```{r birth, message=FALSE, fig.cap="The plot for Life expectancy and Health expenditure by year", fig.align="center"}
# plot Health expenditure and Life expectancy at birth 
plot1 <- ggplot(life_health_expenditure, 
            aes(`Health expenditure per capita(US$)`, 
                `Life expectancy at birth (years)`, 
                size = `Life expectancy at birth (years)`,
                colour = continent)) +
  geom_point(alpha = 0.6, 
             na.rm = TRUE) +
  scale_color_brewer(palette = "Set2")+
  scale_size(range = c(1, 4)) +
  transition_time(Year) +
  #scale_x_continuous(limits = c(0,10000), breaks = 1000)+
  labs(title = 'Year: {frame_time}', 
       x = "Health expenditure per capita(US$)", 
       y = "Life expectancy at birth(years)") +
  ease_aes()+
  theme_bw()

plot1
```


```{r age60, message=FALSE}
# plot Health expenditure and Life expectancy at age60
plot2 <- ggplot(life_health_expenditure, 
            aes(`Health expenditure per capita(US$)` , 
                `Life expectancy at age 60 (years)`, 
                size = `Life expectancy at birth (years)`,
                colour = continent)) +
  geom_point(alpha = 0.6,na.rm = TRUE) +
  scale_color_brewer(palette = "Set2")+
  scale_size(range = c(1, 4)) +
  transition_time(Year) +
  labs(title = 'Year: {frame_time}', 
       x = "Health expenditure per capita(US$)", 
       y = "Life expectancy at age 60") +
  ease_aes()+
  theme_bw()

plot2
```


Figure (\@ref(fig:birth), \@ref(fig:age60)) shows the dynamic relationship between life expectancy at birth(at age 60) and health expenditure from 2000 to 2016. And we find that the countries with the lowest life expectancy and health expenditure are mainly in Africa, while Europe, Oceania and some countries in the Americas are far ahead. Life expectancy in most Asian and American countries is concentrated between 65 and 75 years. Also, almost all countries have increased their spending on health over time and improved life expectancy over time.  
However, there was an outlier country in the Americas showed a significant drop in life expectancy in 2010. Combined with Table \@ref(tab:min_expectancy), we can find that this country is Haiti, which was hit by a large-scale earthquake in 2010 and killed over 230,000 people[@bilham2010lessons], which led to a significant drop in life expectancy in Haiti. Another interesting fact is that the country with the highest health expenditure is not the country with the highest life expectancy. Combine with Table \@ref(tab:min_expectancy), and we can find that the country with the highest life expectancy is in Asia, Japan. Japan has the longest life expectancy in the world mainly because of its government regulation and support for health care[@ikeda2011has].


```{r min_expectancy}
# select the lowest life expectancy to explain what happened in 2010 about this country
Americas_min <- life_health_expenditure %>% 
  filter(continent == "Americas") %>% 
  select(-3,-c(6:8)) %>% 
  slice(which.min(`Life expectancy at birth (years)`)) %>% 
  
  kable(caption = "The lowest life expectancy in the Americas") %>% 
  kable_styling(bootstrap_options = 
                  c("striped", "condensed"), 
                  full_width = F, 
                  position = "center") 

Americas_min

```


```{r max_expectancy}
# select the highest life expectancy
Asia_max <- life_health_expenditure %>% 
  filter(continent == "Asia") %>% 
  select(-3,-c(6:8)) %>% 
  slice(which.max(`Life expectancy at birth (years)`)) %>% 
  
  kable(caption = "The highest life expectancy in the Asia") %>% 
  kable_styling(bootstrap_options = 
                  c("striped", "condensed"), 
                  full_width = F, 
                  position = "center") 

Asia_max

```

## Map for average health expenditure


```{r map_world}
#read world map
map_world <- map_data("world")

```

```{r avg_he}
# compute average health expenditure per capita
avg_health_expenditure <- life_health_expenditure %>% 
  filter(!is.na(`Health expenditure per capita(US$)`)) %>% 
  group_by(Country) %>% 
  summarise(average = mean(`Health expenditure per capita(US$)`)) %>% 
  # modify countries's name to match map data
  mutate(Country = recode(Country, 
    "United States of America" = "USA",
    "Antigua and Barbuda" = "Antigua",
    "Bolivia (Plurinational State of)" = "Bolivia",
    "Brunei Darussalam" = "Brunei",
    "Cabo Verde" = "Cape Verde",
    "Congo" = "Democratic Republic of the Congo",
    "Iran (Islamic Republic of)" = "Iran",
    "Lao People's Democratic Republic" = "Laos",
    "Micronesia (Federated States of)" = "Micronesia",
    "Republic of Korea" = "South Korea",
    "Republic of Moldova" = "Moldova",
    "Republic of North Macedonia" = "Macedonia",
    "Russian Federation" = "Russia",
    "Saint Vincent and the Grenadines" = "Saint Vincent",
    "Syrian Arab Republic" = "Syria",
    "Trinidad and Tobago" = "Trinidad",
    "United Kingdom of Great Britain and Northern Ireland" = "UK",
    "United Republic of Tanzania" = "Tanzania",
    "Venezuela (Bolivarian Republic of)" = "Venezuela",
    "Viet Nam" = "Vietnam"
  ))
  
```


```{r map_he}
#join map_world and health expenditure data
map_health_expenditure <- 
  left_join(avg_health_expenditure, map_world, by = c("Country" = "region")) 

```


```{r map_avg, fig.cap="The average health expenditure per capita(US$) by country", fig.align="center"}
# map average health expenditure
ggplot(map_health_expenditure,
       aes(x = long, 
           y = lat, 
           group = group, 
           fill = average)
       ) +
  
  geom_polygon(color = "#CCCCCC") + 
  scale_fill_gradient(name = "per capita(US$)",
                      low="white", 
                      high="steelblue",
                      limits = c(0, 10000)
                      )+
  ggtitle("The average health expenditure per by country")+
  theme_map()
  
```

By mapping Figure \@ref(fig:map_avg), it can be found that the highest per capita health expenditure is mainly in European countries, the United States, Canada, Australia, New Zealand and Japan. Some countries are 0 due to missing health expenditure data.


## Model for Healthy life expectancy and Health expenditure by continent

```{r lm_data}
# tidy and make lm data
life_health_lm_data <- life_health_expenditure %>% 
  select(Year,continent,Country,-Gender,-c(4:5),c(6:7),9) %>% 
  filter(!is.na(`Health expenditure per capita(US$)`)) %>% 
  filter(!is.na(`Healthy life expectancy (HALE) at birth (years)`)) %>% 
  filter(!is.na(`Healthy life expectancy (HALE) at age 60 (years)`)) %>% 
  pivot_longer(cols = c(4:6), names_to = "Type", values_to = "value") %>% 
  group_by(Year, Type, continent) %>% 
    summarize_at(vars(value), 
               list(Vmean = mean), 
               na.rm = TRUE) %>% 
  pivot_wider(values_from = Vmean, names_from = Type)
  
```


```{r birth_model}
# model: HALE birth and health expenditure
lm_birth <- lm (`Healthy life expectancy (HALE) at birth (years)` ~ `Health expenditure per capita(US$)`, life_health_lm_data)

```


```{r aug_birth_model}
hale_birth_aug <- augment(lm_birth, life_health_lm_data)

```

```{r plot_birth, fig.cap="The plot for Healthy life expectancy at birth(years) and Health expenditure", fig.align="center"}
# plot HALE at birth and health expenditure
ggplot(hale_birth_aug, 
       aes(x = `Health.expenditure.per.capita.US..`, 
           y = .fitted)) + 
  geom_line(colour = "blue") + 
  geom_point(aes(x = `Health.expenditure.per.capita.US..`,
                 y = `Healthy.life.expectancy..HALE..at.birth..years.`,
                 color = continent)) +
  labs(x = "Health Expenditure(US$)", y = "HALE at birth(years).fitted")+
  scale_x_log10() +
  facet_wrap(~Year)

```




```{r age60_model}
# model: HALE at age60 and health expenditure
lm_age60 <- lm (`Healthy life expectancy (HALE) at age 60 (years)` ~ `Health expenditure per capita(US$)`, life_health_lm_data)

```

```{r aug_age60_model}
hale_age60_aug <- augment(lm_age60, life_health_lm_data)

```

```{r plot_age60, fig.cap="The plot for Healthy life expectancy at age 60(years) and Health expenditure", fig.align="center"}
# plot HALE at age60 and health expenditure
ggplot(hale_age60_aug, 
       aes(x = `Health.expenditure.per.capita.US..`, 
           y = .fitted)) + 
  geom_line(colour = "blue") + 
  geom_point(aes(x = `Health.expenditure.per.capita.US..`,
                 y = `Healthy.life.expectancy..HALE..at.age.60..years.`,
                 color = continent)) +
  labs(x = "Health Expenditure(US$)", y = "HALE at age 60 (years).fitted")+
  scale_x_log10() +
  facet_wrap(~Year)


```


From Figure (\@ref(fig:plot_birth), \@ref(fig:plot_age60)), we can see that Africa is the worst fitting one both in Model(at_birth) and Model(at_age60), while Oceania has better fitting in both. As shown in Table \@ref(tab:model_info), we infer that the Model(AT_age60) is better than Model(at_birth), because it has the higher r.squared which explains the variance in the model. Model(at_age60) explains 69.11% of the dependent variable. Also, Model(at_age60) has the lower sigma, which gives it more confidence. Finally, Model(at_age60) has the lower AIC and BIC, which makes it the best model fit[@kenny2015measuring].  

Through Table \@ref(tab:model_ci), we can get the formula of Model(at_birth) as follows:
$$\large{\hat{LifeExp(at~birth)}=57.52+0.00539Health~expenditure}$$

And the formula of Model(at_age60) can be represented by:
$$\large{\hat{LifeExp(at ~age ~60)}=13.20+0.00172Health~expenditure}$$

From the above analysis, we can learn that the amount of health expenditure per capita has a more significant impact on healthy life expectancy at age 60 than healthy life expectancy at birth. It's possibly because people are more likely to get sick as they get older and need more health care to live longer.  


```{r model_info}
bind_rows(
  `Model(at_birth)` = glance(lm_birth),
  `Model(at_age60)` = glance(lm_age60),
  .id = "type") %>% 
  kable(caption = "Model information") %>% 
  kable_styling(bootstrap_options = 
                  c("striped", "condensed"), 
                  full_width = F, 
                  position = "center") 
  
```



```{r model_ci}
bind_rows(
  `Model(at_birth)` = tidy(lm_birth),
  `Model(at_age60)`= tidy(lm_age60),
  .id = "type") %>% 
  kable(caption = "Confidence interval for model") %>% 
  kable_styling(bootstrap_options = 
                  c("striped", "condensed"), 
                  full_width = F, 
                  position = "center") 
  
```





\section*{Life expectancy at birth vs at 60}
Life expectancy is a measure of population longevity which indicates how long a person is expected to live (@tosato2007aging). It can be measured with different levels, @rabbi2013imbalance raises the point that life expectancy at birth would be higher than at any particular age. In this section, life expectancy at birth and at the age of 60 are examined to depict the changes within these indicators across the world from 2000 to 2016. 
```{r tranform}
expectancy_clean <- expectancy %>%
  pivot_longer(cols = `Life expectancy at birth (years)_Both sexes`:`Healthy life expectancy (HALE) at age 60 (years)_2_Female`,
               names_to = c("Type", "Gender"),
               names_sep = "_",
               values_to = "Expectancy") %>%
  mutate(Gender = if_else(Gender == "1", "Male",
                  if_else(Gender == "2", "Female", "Both"))) %>%
  rename(Country = X1_Country,
         Year = X2_Year) %>%
  mutate(Expectancy = as.numeric(Expectancy)) %>%
  mutate(Year = as.numeric(Year))
```
Figure \@ref(fig:3box) presents the overall distribution for life expectancy at birth (years) and life expectancy at age 60 (years) from 2000 to 2016. On the x-axis, Types consist of life expectancy at birth (years), which is positioned as the left boxplot in each sub-graph, and life expectancy at age 60 (years), being the one on the right. The y-axis provides the scale of life expectancy, measured in years. From 2000 to 2016, there is a slight upward-lifting trend for both types of boxplots, implying that people at birth and at the age of 60 are expected to live longer with the progress in time.

```{r 3box, fig.cap="Life expectancy at birth (years) and at age 60 (years) boxplot in 2000, 2004,2008, 2012, 2016"}
box <- expectancy_clean %>%
  na.omit(expectancy_clean) %>%
  filter(Year %in% c(2000, 2004, 2008, 2012, 2016),
         Gender == "Both",
         Type %in% c("Life expectancy at birth (years)", "Life expectancy at age 60 (years)")) %>%
  ggplot(aes(y = Expectancy,
             x = Type,
             color = Year)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  facet_wrap(~ Year, nrow = 1)
box
```


```{r,include=FALSE}
expectancy_clean %>%
  distinct(Country)
```


```{r, message=FALSE}
le_birth_bs <- expectancy_clean %>%
  na.omit(expectancy_clean) %>%
  filter(Type == "Life expectancy at birth (years)",
         Gender == "Both") %>%
  group_by(Year) %>%
  summarise(`average LE at birth`= mean(Expectancy)) 
```

```{r}
meet_average_birth <- expectancy_clean %>%
  na.omit(expectancy_clean) %>%
  filter(Type == "Life expectancy at birth (years)",
         Gender == "Both") %>%
  group_by(Year) %>%
  mutate(`meet average` = if_else(
    condition = Expectancy > (le_birth_bs$`average LE at birth`),
    true = "met",
    false = "unmet"
  )) %>%
  select(Country, Year, Expectancy, `meet average`) 

```

```{r}
meet_count_birth <- meet_average_birth %>%
  count(`meet average`) %>%
  rename(number_countries_birth = n) %>%
  rename(meet_avg_birth = `meet average`)
```


```{r, message=FALSE}
le_60_bs <- expectancy_clean %>%
  na.omit(expectancy_clean) %>%
  filter(Type == "Life expectancy at age 60 (years)",
         Gender == "Both") %>%
  group_by(Year) %>%
  summarise(`average LE at 60` = mean(Expectancy))
```
Table \@ref(tab:3avg_yeari) lists the average life expectancy at birth and at 60, taking into account of all countries where their life expectancy informations are available. Yearly increments at birth and at 60 represent the increasing proportion of the average each year comparing with the average of the former year. Similar to the results of boxplots, the averages had continued increasing since 2001. However, the extent of the rising pattern for average life expectancy at 60 is relatively insignificant to that at birth.
```{r}
avg_yeari <- le_birth_bs %>%
  left_join(le_60_bs, by = "Year") %>%
  mutate(`lag average LE at birth` = lag(`average LE at birth`),
         `Yearly increment at birth` = `average LE at birth` - `lag average LE at birth`,
         `lag average LE at 60` = lag(`average LE at 60`),
         `Yearly increment at 60` = `average LE at 60` - `lag average LE at 60`) %>%
  select(Year, `average LE at birth`, `average LE at 60`, 
         `Yearly increment at birth`, `Yearly increment at 60`) 
```

```{r 3avg_yeari}
avg_yeari %>%
  kable(caption = "Average life expectancy at birth, at age 60 (years) and their yearly increment")
```
Figure \@ref(fig:3avg_yeari_gr) exhibits the yearly improvements in average life expectancy both at birth and at 60 from 2000 to 2016. The increment for life expectancy at 60 was lower than that at birth for all times from 2000 to 2016. Nevertheless, the patterns for both of them are similar. For example, their increments in 2016 nearly drop to zero. On the next year, they both increase to the maximum increment from 2000 to 2016, as shown as the peaks. 
```{r 3avg_yeari_gr, fig.cap="Yearly increment of average life expectancy at birth and at age 60 (years)"}
avg_yeari_gr <- avg_yeari %>%
  pivot_longer(., cols = c(`Yearly increment at birth`, `Yearly increment at 60`),
               names_to = "LE_Type", values_to = "Increment") %>%
  ggplot(aes(x = Year, y = Increment, color = LE_Type)) +
  geom_point() +
  geom_line() 
avg_yeari_gr
```

```{r}
meet_average_60 <- expectancy_clean %>%
  na.omit(expectancy_clean) %>%
  filter(Type == "Life expectancy at age 60 (years)",
         Gender == "Both") %>%
  group_by(Year) %>%
  mutate(`meet average` = if_else(
    condition = Expectancy > (le_60_bs$`average LE at 60`),
    true = "met",
    false = "unmet"
  )) %>%
  select(Country, Year, Expectancy, `meet average`) 
```

```{r}
meet_count_60 <- meet_average_60 %>%
  count(`meet average`) %>%
  rename(number_countries_60 = n) %>%
  rename(meet_avg_60 = `meet average`)
```
Table \@ref(tab:3count_birth_60) contains the number of countries which had and had not met the average life expectancy from 2000 to 2016.
```{r, message=FALSE}
count_birth_60 <- cbind(meet_count_birth, meet_count_60) %>%
  select(-Year...4, -meet_avg_60) %>%
  rename(Year = Year...1, meet_avg = meet_avg_birth) 
  
```

```{r 3count_birth_60}
count_birth_60 %>%
kable(caption = "Number of countries met and unmet average life expectancy at birth, at age 60 (years) each year")
```

```{r}
count_met_graph <- count_birth_60 %>%
  filter(meet_avg == "met") %>%
  pivot_longer(., cols = c(number_countries_birth, number_countries_60),
               names_to = "LE_count_type", values_to = "number_met_countries") %>%
  ggplot(aes(x = Year, y = number_met_countries, color = LE_count_type)) +
  geom_point() +
  geom_line() 
```

```{r}
count_unmet_graph <- count_birth_60 %>%
  filter(meet_avg == "unmet") %>%
  pivot_longer(., cols = c(number_countries_birth, number_countries_60),
               names_to = "LE_count_type", values_to = "number_unmet_countries") %>%
  ggplot(aes(x = Year, y = number_unmet_countries, color = LE_count_type)) +
  geom_point() +
  geom_line() 
```
Figure \@ref(fig:3combine_count) combines the number of countries that had met the average life expectancy at birth, at 60 and number of countries that had not met the averages from 2000 to 2016. As years pass on, more countries had been meeting and exceeding the average life expectancy at birth and at 60 each year. However, a pheonomenon is revealed here - Although the number of countries meeting average life expectancy at 60 is lower than that at birth, the number of countries which did not meet the average is much higher. Further, since 2001, the number of countries meeting average life expectancy at birth had already surpassed that of not meeting. Yet for life expectancy at 60, it was until 2013, the number of countries meeting the average life expectancy at 60 began to exceed that of unmet number of countries.
```{r 3combine_count, fig.cap="Trend of number of countries meeting and not meeting each year average LE at birth and 60", warning=FALSE}
library(gridExtra)
combine_count <- grid.arrange(count_met_graph, count_unmet_graph)
```



\section*{Life expectancy and Gross Domestic Product}

